### 2020-03-23 revision.Rmd
Submitted Revision 1

---
title             : "Quantitative involvement moderates longitudinal associations between mothers' and fathers' qualitative parenting and children's emotional adjustment"
shorttitle        : "INVOLVEMENT AND EMOTIONAL ADJUSTMENT"

@@

abstract: |
  The question of whether fathers and mothers play similar roles in children’s
  emotional adjustment has been addressed using either qualitative parenting 
  measures or  quantitative child care involvement. This
  longitudinal study (N = 1884, age 6-12) bridges these approaches, 
  investigating qualitative and quantitative involvement, and their interaction,
  in relation to emotional adjustment. On average, evidence for mothers'
  and fathers' unique roles was limited.
  When both parents were more involved, both engaged
  more in desirable qualitative parenting. When mothers were disproportionately
  involved, however, parenting roles were more differentiated:
  Fathers displayed less
  desirable qualitative parenting, and children were less well-adjusted.
  This suggests that all family members benefit from a more equal
  childcare distribution.

keywords          : "involvement; emotional adjustment; fathers; parenting; longitudinal"
wordcount         : "7356"

@@

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their children). Nevertheless, the existing literature has two major shortcomings. Firstly, in assessing linkages with children’s emotional adjustment, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. Children’s emotional adjustment might, for example, be differently impacted by a parent's warmth if that parent is generally highly involved with the day-to-day minutiae of childrearing, than when (s)he is generally absent. Conversely, children’s emotional adjustment might elicit different qualitative parenting behaviors from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be sanctioned. The present study set out to address this question.

Secondly, although fathers have been theorized to play a unique role in emotional adjustment development, empirical work has rarely addressed the joint and relative contributions of fathers' and mothers' involvement. The present study overcomes these limitations by incorporating both quantitative and quantitative measures of mothers' and fathers' involvement. We use sophisticated methods to examine how joint and relative quantitative involvement moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

## Fatherhood theory

Attachment relationships are grounded in a balance between security and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Fatherhood researchers have posited that warmth and security are more central to mother-child attachment relationships, whereas father-child relationships have been construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. Consequently, fathers and mothers are theorized to influence emotional adjustment development in somewhat different ways.

The past decades have seen major strides in the consolidation of theories to guide fatherhood research. Early fatherhood research primarily operationalized involvement in *quantitative* terms: Presence versus absence in divorced families, and the sharing of childcare duties in dual-earner families [@lambBiosocialPerspectivePaternal1987]. Over time, the notion the fathers should share child-rearing responsibilities with mothers, adopting roles historically fulfilled by mothers, gained ground [@pleckFatherhoodIdealsUnited1997]. Consequently, definitions of father involvement evolved to incorporate *qualitative* dimensions, such as warmth and control [@maccobySocializationContextFamily1983; @pleckPaternalInvolvementRevised2004].

It is now widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004]. Nevertheless, these dimensions are rarely investigated together, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors have been operationalized in terms of the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007]. Therefore we hypothesized that parental warmth would be positively associated with children's emotional adjustment, $H_{1a}$.

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Some of these might help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. Therefore we hypothesized that consistency would be positively associated with emotional adjustment, $H_{1b}$. There is less empirical evidence for linkages between inductive reasoning and children’s emotional adjustment. Given that inductive inductive reasoning promotes the internalization of parental expectations, including display rules for emotions [@grusecImpactParentalDiscipline1994], we hypothesized that it would also be positively related to emotional adjustment, $H_{1c}$.

Negative parenting, which we operationalized as parental anger, is an additional qualitative parenting dimension that is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. 
Parental anger expressions negatively impact children's emotional adjustment [@scaramellaParentingChildrenDistress2008], because parental anger is upsetting, and because anger reflects poor parental emotion regulation, which serves as a model for children [@changHarshParentingRelation2003]. We thus hypothesized that it would be negatively related to emotional adjustment, $H_{1d}$.

### Parent or child effects?
<!-- Hier iets doen 
Was wel echt meer control, maar rigid control oid?-->
A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that warmth, consistency, and inductive reasoning positively predict, and anger negatively predicts, emotional adjustment, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.
<!-- Neppl 2009: Adolescents' harsh behavior towards parents predicts harsh parenting, and positive behavior predicts positive parenting -->

@@

Several studies have re-examined the direction of effects between parents and children using these methods, but (to our knowledge) only in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods, disappeared when stable between-family differences in parenting and adjustment were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that most associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2018a]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample. These two studies also illustrate that applying these new methods changes the pattern of results in unexpected ways. Consequently, it is impossible to make specific predictions about the pattern of between-family correlations and within-family predictive effects. For now, we will assume that $H_{1a-d}$ and $H_{2a-d}$ apply at both the between- and within-family level.
<!-- Hier iets over de hypotheses; dat we niet weten wat er within en wat between speelt 
We are, however, a bit cautious about formulating more focused hypotheses about within- versus between-family effects than the one formulated above. Because there is no precedent for using the random-intercept cross-lagged panel model or its multilevel DSEM extension in the literature on parental involvement and emotional adjustment, there is no literature upon which to base such hypotheses. Our examination of between and within-family effects will therefore mainly be exploratory. 
-->

@@

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' qualitative involvement [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. Some have theorized that mothers and fathers might fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage and respond supportively to emotional expression to a greater extent than fathers, whereas fathers structure and sanction emotion expressions more than mothers [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, for adolescents' emotional adjustment, maternal warmth was found to be a more salient socialization mechanism than demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demanding behaviors, such as consistency and inductive reasoning, $H_{3}$.

@@

Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the **division** of childcare tasks is more relevant for children's adjustment [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Effects that appear to be due to fathers' greater involvement might be due to children receiving **more** total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; @keizerInfluenceFathersMothers2019].

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. We therefore hypothesize that greater joint care should positively predict children's emotional adjustment, $H_{6a}$. By contrast, we predict that greater relative involvement will negatively predict emotional adjustment, $H_{6b}$, because children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001].

@@

Child sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls than boys, and more anger towards boys than girls, $H_{7a-d}$. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent, $H_{8a-d}$.

Furthermore, although sex differences in emotional adjustment are sometimes found in older children [@silkAdolescentsEmotionRegulation2003; @vanlissaRoleFathersMothers2018a], a meta-analysis found little evidence for such sex differences in young children [@else-questGenderDifferencesTemperament2006a]. Consequently, we have no hypothesis regarding sex differences in children's average levels of emotional adjustment.

@@

All analyses were conducted using Mplus 8.0 and the R-package MplusAutomation. This type of multilevel autoregressive model can only be estimated using Bayesian methods. We used two Gibbs-sampler chains of 50000 iterations each. Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing at random. The missing data handling is likelihood based and guarantees consistent estimation as long as the missing data is missing at random (MAR). We applied mildly informative inverse Wishart priors to the variance/covariance matrices, and normal priors centered on zero for all regression coefficients. We assessed convergence through visual inspections of the chains and posterior distributions. We report the posterior median, and a 95% highest posterior density interval (HPDI). This is the "narrowest" possible interval that contains 95% of samples from the posterior distribution. It represents the most likely population values of the parameter in question. Results are best interpreted by examining the effect size (standardized regression coefficients and correlations), and the width of the HPDI. Significance is not a Bayesian concept, but we use the shorthand "significant" to refer to estimates for which less than 5% of the posterior distribution had the opposite sign from the estimate. All standardized and unstandardized model parameters are reported in Supplementary Table 4.

@@

We hypothesized that warmth, consistency, and inductive reasoning would be positively associated, and anger would be negatively associated, with emotional adjustment. At the between-family level, we examined correlations between the random intercepts of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). Random intercepts reflect the average level of each variable across all waves of the study. Counter to hypothesis $H_{1a}$, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypothesis $H_{1b}$, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypothesis $H_{1c}$, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. This means that, in families where parents use more inductive reasoning, children displayed poorer emotional adjustment. Finally, as hypothesized in $H_{1d}$, we found significant negative correlations between maternal and paternal anger, and emotional adjustment.

@@

We also examined within-family correlations, which are based on the residuals of the study variables, across all waves and families (Table \@ref(tab:tabcors)). In support of hypothesis $H_{1a}$, we found significant positive within-family level associations between emotional adjustment and fathers' and mothers' warmth. In partial support of hypotheses $H_{1b,c}$, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to hypothesis $H_{1d}$, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger. Counter to hypothesis $H_{3}$, none of the differences between mothers and fathers in the strength of either between- or within-family level correlations were significant (see Supplementary Table S5). This includes the within-family correlation between warmth and emotional adjustment, whose effect size was nearly twice as large for fathers as compared to mothers.

@@

At the within-family level, we examined over-time predictive effects between qualitative parenting and emotional adjustment. The intercepts of these random effects reflect the average standardized regression coefficient across families. The variances in Table \@ref(tab:tabcrosspath) indicate whether this coefficient differs in strength (and sign) between families. We report unstandardized variances, because the standardized variances are, by definition, set to 1. With regard to within-families parenting effects, we found no evidence for hypotheses $H_{1b-d}$, as none of the intercepts differed significantly from zero. In line with hypothesis $H_{1a}$, however, on average, a maternal warmth positively predicted children's emotional adjustment in the next wave. The standardized regression coefficient of this effect, and others, is interpreted as follows: A $.99$ standard deviation increase in children's emotional adjustment in the next wave (see Table \@ref(tab:tabcrosspath)). The variances of all parenting effects were significant, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were all significantly larger for fathers than for mothers (see Supplementary Table S5), indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.

>>>>

---
title             : "Quantitative involvement moderates the within-family interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment"
shorttitle        : "INVOLVEMENT AND EMOTIONAL ADJUSTMENT"

@@

abstract: |
  Ample research has shown that children's emotional adjustment and parenting
  are related. Several important questions remain unaddressed. First, both
  quantitative involvement (time spent on childcare), and qualitative
  involvement (specific parenting behaviors) have been implicated in emotional
  adjustment development. Yet these involvement dimensions are rarely
  investigated jointly or in interaction. Secondly, although fathers have been
  theorized to contribute uniquely to emotional adjustment development, they
  are not routinely included in research. Consequently, the potential
  implications of joint- (total) and relative (who does what) quantitative
  involvement for the interplay between qualitative parenting and emotional
  adjustment have not been adequately researched. This longitudinal study of
  Australian families (N = 1884, from age 6-12) addressed these questions,
  using dynamic structural equation modeling to examine the within-family
  interplay between qualitative parenting and emotion regulation, while
  controlling for stable between-family differences in these constructs. Joint
  and relative quantitative involvement, and child sex, were included as
  between-family moderators of within-family processes. We primarily found
  between-family correlations between qualitative involvement and emotional
  adjustment. Evidence for within-family effects was limited: On average,
  maternal warmth positively predicted emotional adjustment. Several child
  effects indicated that emotional adjustment evokes changes in parenting.
  Within-family effects varied substantially across families however,
  especially for fathers. When joint quantitative involvement was higher, both
  parents engaged in more desirable qualitative parenting. When mothers were
  relatively more involved than fathers, parental influence was more
  differentiated, fathers displayed less desirable qualitative parenting, and
  children were less well-adjusted. This suggests that all family members
  benefit from a more equal childcare distribution.

keywords          : "involvement; emotional adjustment; fathers; parenting; longitudinal"
wordcount         : "8634"

@@

\vspace{\baselineskip}
Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children’s emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children’s poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

Secondly, although emotional adjustment is one particular area of development where fathers have been theorized to play a unique role [@paquetteTheorizingFatherchildRelationship2004], fathers are still not consistently included in research [@faganShouldResearchersConceptualize2014]. Consequently, prior research has not sufficiently addressed mothers' and fathers' joint and unique contributions to emotional adjustment development. The present study overcomes this limitation by incorporating both parents. We use sophisticated methods to examine how joint and relative quantitative involvement at age 6 moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors are related to the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's emotional adjustment [@morrisRoleFamilyContext2007]. Therefore we hypothesized that parental warmth would be positively associated with children's emotional adjustment, $H_{1a}$.<!--@khalequePerceivedParentalAcceptanceRejection2002; -->

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses parenting behaviors related to setting and enforcing expectations for children. Some of these behaviors help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency correlates positively with emotional adjustment [@brandPerceivedParentalInconsistency1990]. Therefore we also hypothesized that consistency would be positively associated with emotional adjustment, $H_{1b}$. There is less empirical evidence for links between inductive reasoning and emotional adjustment. Given that inductive reasoning promotes the internalization of parental expectations, including display rules for emotions [@grusecImpactParentalDiscipline1994], we hypothesized that it would also be positively related to emotional adjustment, $H_{1c}$.

Negative parenting, which we operationalized as parental anger, is a qualitative parenting dimension not originally subsumed in the parenting styles typology. It is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. Parental anger negatively impacts children's emotional adjustment [@scaramellaParentingChildrenDistress2008], because it is upsetting, and because anger reflects poor parental emotion regulation, which serves as a model for children [@changHarshParentingRelation2003]. We thus hypothesized that anger would be negatively related to emotional adjustment, $H_{1d}$.

### Parent or child effects?

A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that emotional adjustment positively predicts warmth, consistency, and inductive reasoning, and negatively predicts anger, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.
<!-- Neppl 2009: Adolescents' harsh behavior towards parents predicts harsh parenting, and positive behavior predicts positive parenting -->

@@

Several studies have re-examined the direction of effects between parents and children using these methods, but (to our knowledge) only in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods, disappeared when stable between-family differences in parenting and adjustment were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that most associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2018a]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample. These two studies also illustrate that applying these new methods changes the pattern of results in unexpected ways. Consequently, it is impossible to make specific predictions about the pattern of between-family correlations and within-family predictive effects. For now, we will assume that $H_{1a-d}$ and $H_{2a-d}$ apply at both the between- and within-family level.

@@

The notion that fathers should share child-rearing responsibilities with mothers has become widely accepted in the Western world [@pleckFatherhoodIdealsUnited1997]. Consequently, researchers have come to study qualitative involvement along the same dimensions for mothers and fathers [@maccobySocializationContextFamily1983; @pleckPaternalInvolvementRevised2004]. Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers [@flouriFathersBehaviorsChildren2010]. Some research suggests that both parents influence children similarly, but that effects might be stronger for mothers [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014]. Others have argued that mothers and fathers might contribute through somewhat different mechanisms. Attachment relationships are grounded in a balance between warmth and security, and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Some have proposed that warmth is more central to mother-child attachment relationships, whereas father-child relationships can be construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000].

In line with this theory, empirical research has found that mothers' responses to emotional expression are more warm and supportive than fathers' responses, which are more demanding and punitive [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. In adolescence, maternal warmth was found to be a more salient socialization mechanism of emotional adjustment than maternal demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is initial evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is somewhat incongruous with the warm "attachment" relationship typically expected of mothers. Based on this research, we therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demandingness, operationalized as consistency and inductive reasoning, $H_{3}$.

@@

Early fatherhood research primarily operationalized involvement in *quantitative* terms: Presence versus absence in divorced families, and the sharing of childcare duties in dual-earner families [@lambBiosocialPerspectivePaternal1987]. Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the *division* of childcare tasks is more relevant for children's adjustment [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Effects that appear to be due to fathers' greater involvement might be due to children receiving *more* total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; @keizerInfluenceFathersMothers2019].

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. We therefore hypothesize that greater joint care should positively predict children's emotional adjustment, $H_{6a}$. Further, it has been argued that children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001]. We therefore predict that greater relative involvement will negatively predict emotional adjustment, $H_{6b}$.

@@

Child sex differences are relevant in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls than boys, and more anger towards boys than girls, $H_{7a-d}$. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent, $H_{8a-d}$. Sex differences in emotional adjustment have been found in older children [@silkAdolescentsEmotionRegulation2003; @vanlissaRoleFathersMothers2018a]. However, a meta-analysis found little evidence for such sex differences in young children [@else-questGenderDifferencesTemperament2006a]. Consequently, we have no hypothesis regarding sex differences in average levels of emotional adjustment.

@@

Analyses were conducted using Mplus 8.0 and the R-package MplusAutomation. This type of multilevel autoregressive model can only be estimated using Bayesian methods. We used two Gibbs-sampler chains of 50000 iterations each. Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing-at-random (a non-parameteric MCAR test resulted in $p > .05$). We used mildly informative inverse Wishart priors for covariance matrices, and normal priors centered on zero for regression coefficients. We assessed convergence through visual inspection of the chains and posterior distributions. We report the posterior median, and a 95% highest posterior density interval (HPDI). This is the "narrowest" possible interval that contains 95% of samples from the posterior distribution. It represents the most likely population values of the parameter in question. Results are best interpreted by examining the effect size (correlations and standardized regression coefficients), and the width of the HPDI. Significance is not a Bayesian concept, but we use the shorthand "significant" to refer to estimates for which less than 5% of the posterior distribution had the opposite sign from the estimate. Supplementary Table S4 reports all standardized and unstandardized model parameters. All code and output are available online at [masked].

@@

We hypothesized that warmth, consistency, and inductive reasoning would be positively associated, and anger would be negatively associated, with emotional adjustment. At the between-family level, we examined correlations between the random intercepts of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). Random intercepts reflect the average level of each variable across all waves of the study. Counter to hypotheses $H_{1,2a}$, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypotheses $H_{1,2b}$, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypotheses $H_{1,2c}$, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. This means that, in families where parents use more inductive reasoning, children displayed poorer emotional adjustment. Finally, as hypothesized in $H_{1,2d}$, we found significant negative correlations between maternal and paternal anger, and emotional adjustment.

@@

We also examined within-family correlations, which are based on the residuals of the study variables, across all waves and families (Table \@ref(tab:tabcors)). In support of hypothesis $H_{1a}$, we found significant positive within-family level associations between emotional adjustment and fathers' and mothers' warmth. In partial support of hypotheses $H_{1b,c}$, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to hypothesis $H_{1d}$, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger. Counter to hypothesis $H_{3}$, none of the differences between mothers and fathers in the strength of either between- or within-family level correlations were significant (see Supplementary Table S4). This includes the within-family correlation between warmth and emotional adjustment, whose effect size was nearly twice as large for fathers as compared to mothers.

@@

At the within-family level, we examined over-time predictive effects between qualitative parenting and emotional adjustment. The intercepts of these random effects reflect the average standardized regression coefficient across families. The variances in Table \@ref(tab:tabcrosspath) indicate whether this coefficient differs in strength (and sign) between families. We report unstandardized variances, because the standardized variances are, by definition, set to 1. With regard to within-families parenting effects, we found no evidence for hypotheses $H_{1b-d}$, as none of the intercepts differed significantly from zero. In line with hypothesis $H_{1a}$, however, on average, a maternal warmth positively predicted children's emotional adjustment in the next wave. The standardized regression coefficient of this effect, and others, is interpreted as follows: A $.99$ standard deviation increase in children's emotional adjustment in the next wave (see Table \@ref(tab:tabcrosspath)). The variances of all parenting effects were significant, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were all significantly larger for fathers than for mothers (see Supplementary Table S4), indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.
